```{r}
#run
```{r}
# ====================================================================
# BETA-DIVERSITY (BRAY-CURTIS) – GROUP (AB, PPI, OTH, NO) + TIMEPOINTS
# WITH ANOSIM ON PCoA AND STANDARD SIGNIFICANCE STARS ON BOXPLOTS
# ====================================================================

library(phyloseq)
library(tidyverse)
library(vegan)
library(ggpubr)
library(patchwork)

# ------------------------------------------------------------
# 1. Load and clean phyloseq object
# ------------------------------------------------------------
ps <- readRDS('./RSF_all_physeq_wgs_COUNTS_v3.rds')
ps@sam_data$Batch <- as.factor(ps@sam_data$Batch)
bad_batch_samples <- c('B-024-1', 'B-024-2', 'B-027-2', 'B-030-2')
if (length(bad_batch_samples) > 0) {
  ps <- prune_samples(!(sample_names(ps) %in% bad_batch_samples), ps)
}

# ------------------------------------------------------------
# 2. Assign Group (AB, PPI, OTH, NO) – directly replace original Group
# ------------------------------------------------------------
sam <- data.frame(sample_data(ps), stringsAsFactors = FALSE)
sam$SampleID <- rownames(sam)

# Build new grouping, overwriting the existing 'Group' column
new_group <- rep(NA_character_, nrow(sam))
new_group[sam$Group == "A" ] <- "AB"
new_group[sam$Group == "C"] <- "NO"
new_group[sam$Proton_Pump_Inhibitors == "Yes" & is.na(new_group)] <- "PPI"
new_group[sam$Group == "B" & is.na(new_group) & sam$Proton_Pump_Inhibitors == "No"] <- "OTH"

sam$Group <- factor(new_group, levels = c("AB", "PPI", "OTH", "NO"))
sample_data(ps) <- sam
group_levels <- levels(sam$Group)
# Keep only samples with defined Group and timepoints -1, -2
ps_sub <- ps
sample_data(ps_sub)$Group <- sam$Group
ps_sub <- subset_samples(ps_sub, !is.na(Group) & Timepoint %in% c("-1", "-2"))
ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)
sample_data(ps_sub)$Timepoint <- factor(sample_data(ps_sub)$Timepoint, levels = c("-1", "-2"))

# ------------------------------------------------------------
# 3. Bray-Curtis distance and PCoA
# ------------------------------------------------------------
ps_rel <- transform_sample_counts(ps_sub, function(x) x / sum(x))
dist_bc <- phyloseq::distance(ps_rel, method = "bray")
pcoa <- ordinate(ps_rel, method = "PCoA", distance = dist_bc)
pcoa_df <- data.frame(sample_data(ps_rel))
pcoa_df$PC1 <- pcoa$vectors[,1]
pcoa_df$PC2 <- pcoa$vectors[,2]
var_explained <- round(pcoa$values$Relative_eig[1:2] * 100, 2)

# ------------------------------------------------------------
# 4. ANOSIM overall (for annotation)
# ------------------------------------------------------------
anosim_res <- anosim(dist_bc, pcoa_df$Group, permutations = 999)
anosim_text <- paste("ANOSIM R =", round(anosim_res$statistic, 3),
                     "   p =", format(anosim_res$signif, scientific = FALSE, digits = 3))

# ------------------------------------------------------------
# 5. PCoA plot with ANOSIM annotation
# ------------------------------------------------------------
p_pcoa <- ggplot(pcoa_df, aes(x = PC1, y = PC2, color = Group, shape = Timepoint)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(aes(group = Group), linetype = 2, size = 0.8) +
  scale_color_manual(values = c("AB" = "#E41A1C", "PPI" = "#377EB8",
                                "OTH" = "#4DAF4A", "NO" = "#984EA3")) +
  theme_bw(base_size = 12) +
  labs(x = paste0("PC1 (", var_explained[1], "%)"),
       y = paste0("PC2 (", var_explained[2], "%)"),
       title = "PCoA – Bray-Curtis distances") +
  theme(legend.position = "bottom") +
  annotate("text", x = -Inf, y = Inf, label = anosim_text,
           hjust = -0.1, vjust = 1.1, size = 3.5, alpha = 0.8)

# ------------------------------------------------------------
# 6. Boxplots for PC1 and PC2 – all pairwise significance brackets
# ------------------------------------------------------------

# Helper: unpaired Wilcoxon test for two groups
calc_wilcox <- function(data, xvar, yvar, g1, g2) {
  vals <- data[[yvar]]
  grp  <- data[[xvar]]
  test <- wilcox.test(vals[grp == g1], vals[grp == g2], exact = FALSE)
  test$p.value
}

group_levels <- levels(pcoa_df$Group)   # "AB" "PPI" "OTH" "NO"

# all unique combinations
pairs_all <- combn(group_levels, 2, simplify = FALSE)   # list of 6 pairs

# compute p-values for all pairs on both PCs
pval_df <- do.call(rbind, lapply(pairs_all, function(p) {
  data.frame(
    Group1 = p[1],
    Group2 = p[2],
    pval_PC1 = calc_wilcox(pcoa_df, "Group", "PC1", p[1], p[2]),
    pval_PC2 = calc_wilcox(pcoa_df, "Group", "PC2", p[1], p[2]),
    stringsAsFactors = FALSE
  )
}))

# function to convert p-value to significance stars
star <- function(p) {
  ifelse(p < 0.001, "***",
         ifelse(p < 0.01, "**",
                ifelse(p < 0.05, "*", "ns")))
}
pval_df$signif_PC1 <- star(pval_df$pval_PC1)
pval_df$signif_PC2 <- star(pval_df$pval_PC2)

# overall Kruskal–Wallis
kw_pc1 <- kruskal.test(PC1 ~ Group, data = pcoa_df)
kw_pc2 <- kruskal.test(PC2 ~ Group, data = pcoa_df)

# x positions for the four groups
group_x <- setNames(seq_along(group_levels), group_levels)

# --------------------------------------------------
# PC1 boxplot
# --------------------------------------------------
y_range_pc1 <- diff(range(pcoa_df$PC1, na.rm = TRUE))
y_max_pc1   <- max(pcoa_df$PC1, na.rm = TRUE)
step_pc1    <- 0.07 * y_range_pc1      # vertical spacing between brackets
bracket_base_pc1 <- y_max_pc1 + 0.05 * y_range_pc1   # first bracket height

p_pc1_box <- ggplot(pcoa_df, aes(x = Group, y = PC1, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) +
  scale_fill_manual(values = c("AB" = "#E41A1C", "PPI" = "#377EB8",
                               "OTH" = "#4DAF4A", "NO" = "#984EA3")) +
  theme_bw(base_size = 12) +
  labs(y = "PC1 score", title = "PC1") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# add pairwise brackets (6 comparisons)
for (i in 1:nrow(pval_df)) {
  g1 <- pval_df$Group1[i]
  g2 <- pval_df$Group2[i]
  x1 <- group_x[g1]
  x2 <- group_x[g2]
  y_val <- bracket_base_pc1 + (i-1) * step_pc1
  p_pc1_box <- p_pc1_box +
    annotate("segment", x = x1, xend = x2, y = y_val, yend = y_val) +
    annotate("text", x = mean(c(x1, x2)), y = y_val + 0.015 * y_range_pc1,
             label = pval_df$signif_PC1[i], size = 4.5, vjust = 0)
}

# add Kruskal–Wallis annotation above all brackets
kw_y_pc1 <- bracket_base_pc1 + (nrow(pval_df) + 0.7) * step_pc1
p_pc1_box <- p_pc1_box +
  annotate("text", x = 2.5, y = kw_y_pc1,
           label = paste0("Kruskal–Wallis p = ", format(kw_pc1$p.value, digits = 2, scientific = TRUE)),
           size = 3.5, hjust = 0.5)

# --------------------------------------------------
# PC2 boxplot (identical logic)
# --------------------------------------------------
y_range_pc2 <- diff(range(pcoa_df$PC2, na.rm = TRUE))
y_max_pc2   <- max(pcoa_df$PC2, na.rm = TRUE)
step_pc2    <- 0.07 * y_range_pc2
bracket_base_pc2 <- y_max_pc2 + 0.05 * y_range_pc2

p_pc2_box <- ggplot(pcoa_df, aes(x = Group, y = PC2, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) +
  scale_fill_manual(values = c("AB" = "#E41A1C", "PPI" = "#377EB8",
                               "OTH" = "#4DAF4A", "NO" = "#984EA3")) +
  theme_bw(base_size = 12) +
  labs(y = "PC2 score", title = "PC2") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

for (i in 1:nrow(pval_df)) {
  g1 <- pval_df$Group1[i]
  g2 <- pval_df$Group2[i]
  x1 <- group_x[g1]
  x2 <- group_x[g2]
  y_val <- bracket_base_pc2 + (i-1) * step_pc2
  p_pc2_box <- p_pc2_box +
    annotate("segment", x = x1, xend = x2, y = y_val, yend = y_val) +
    annotate("text", x = mean(c(x1, x2)), y = y_val + 0.015 * y_range_pc2,
             label = pval_df$signif_PC2[i], size = 4.5, vjust = 0)
}

kw_y_pc2 <- bracket_base_pc2 + (nrow(pval_df) + 0.7) * step_pc2
p_pc2_box <- p_pc2_box +
  annotate("text", x = 2.5, y = kw_y_pc2,
           label = paste0("Kruskal–Wallis p = ", format(kw_pc2$p.value, digits = 2, scientific = TRUE)),
           size = 3.5, hjust = 0.5)

# ------------------------------------------------------------
# Combine plots (this line stays as before)
# ------------------------------------------------------------
combined_plot <- p_pcoa / (p_pc1_box | p_pc2_box) + 
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(face = "bold"))

# ------------------------------------------------------------
# 7. ANOSIM pairwise (optional) – output to console
# ------------------------------------------------------------
pairwise_anosim <- function(dist, grouping) {
  groups <- unique(grouping)
  res <- data.frame()
  for (i in 1:(length(groups)-1)) {
    for (j in (i+1):length(groups)) {
      keep <- grouping %in% c(groups[i], groups[j])
      sub_dist <- as.dist(as.matrix(dist)[keep, keep])
      sub_group <- grouping[keep]
      test <- anosim(sub_dist, sub_group, permutations = 999)
      res <- rbind(res, data.frame(Group1 = groups[i], Group2 = groups[j],
                                   R = test$statistic, p.value = test$signif))
    }
  }
  return(res)
}
pairwise <- pairwise_anosim(dist_bc, pcoa_df$Group)
cat("\n=== Pairwise ANOSIM (Group) ===\n")
print(pairwise)

# ------------------------------------------------------------
# 8. Save outputs
# ------------------------------------------------------------
out_dir <- "/beta_diversity_group_analysis"
dir.create(out_dir, showWarnings = FALSE)

ggsave(file.path(out_dir, "PCoA_with_boxplots.png"), combined_plot,
       width = 10, height = 10, dpi = 300)
ggsave(file.path(out_dir, "PCoA_ordination.png"), p_pcoa,
       width = 7, height = 5, dpi = 300)
ggsave(file.path(out_dir, "PC1_boxplot.png"), p_pc1_box,
       width = 5, height = 5, dpi = 300)
ggsave(file.path(out_dir, "PC2_boxplot.png"), p_pc2_box,
       width = 5, height = 5, dpi = 300)
write.csv(pairwise, file.path(out_dir, "pairwise_anosim_group.csv"), row.names = FALSE)

cat("\nAll results saved in:", out_dir, "\n")

```
