#libs
```{r}
library(phyloseq)
library(tidyverse)
library(tibble)
```

# metadata table
```{r}
meta <- read.csv('/home/sonec_iv/RSF_res_4_ps/metadata_filtered_eng.txt')
rownames(meta) <- meta$No_Patient
#meta$ID <- gsub('MPAR','MPAR_',meta$ID)
meta <- as.data.frame(sapply(meta, function(x) replace(x, x == "", "No")))
colnames(meta) <- gsub(" ", "_", colnames(meta))
names(meta)[duplicated(names(meta))]
meta$Age = as.numeric(meta$Age)
meta <- meta %>%
  mutate(Age = as.numeric(Age),  # Convert age to numeric
         Age_group = case_when(
           between(Age, 19, 35) ~ '19_35',
           between(Age, 36, 45) ~ '36_45',
           between(Age, 46, 55) ~ '46_55',
           between(Age, 56, 65) ~ '56_65',
           is.na(Age) ~ 'no_data',  
           TRUE ~ 'above_65'
         ))
meta$Weight <- as.numeric(meta$Weight)
meta$Height <- as.numeric(meta$Height)
#numeric_cols <- sapply(meta, is.numeric)
#meta[numeric_cols] <- lapply(meta[numeric_cols], as.factor)
calculate_bmi <- function(weight, height) {
  if (is.na(weight) || is.na(height)) {
    return(NA)  # Skip calculation if either value is NA
  } else {
    bmi <- weight / ((height/100)^2)  # Calculate BMI only if both values are provided
    return(bmi)
  }
}
meta$BMI <- mapply(calculate_bmi, as.numeric(meta$Weight), as.numeric(meta$Height))
meta <- meta %>%
  mutate(BMI = as.numeric(BMI),  # Convert age to numeric
         BMI_group = case_when(
           between(BMI, 18.5, 25.0) ~ 'normal',
           between(BMI, 25.00001, 30.0) ~ 'overweight',
           between(BMI, 30.00001, 99) ~ 'obesity',
           between(BMI, 0, 18.49999) ~ 'underweight',
           is.na(BMI) ~ 'no_data',  
           TRUE ~ 'no_data'
         ))
#write_csv(meta, 'metadata_eng_prepared.csv')
```         

#duplicate sample info so metadata would have 2 timepoints
```{r}
# FIX: Create expanded metadata with proper indexing
expanded_meta <- data.frame()

# Loop through each ROW (by index), not by patient ID
for (i in 1:nrow(meta)) {
  # Get the patient row
  patient_row <- meta[i, , drop = FALSE]
  patient_id <- patient_row$No_Patient
  
  # Create timepoint 1
  time1 <- patient_row
  time1$SampleID <- paste0(patient_id, "-1")
  time1$Timepoint <- "-1"
  
  # Create timepoint 2
  time2 <- patient_row
  time2$SampleID <- paste0(patient_id, "-2")
  time2$Timepoint <- "-2"
  
  # Add to expanded metadata
  expanded_meta <- rbind(expanded_meta, time1, time2)
}

# Set row names to SampleID
rownames(expanded_meta) <- expanded_meta$SampleID

# Verify
cat("\n=== VERIFICATION ===\n")
cat("Number of original patients:", nrow(meta), "\n")
cat("Number of expanded samples:", nrow(expanded_meta), "\n")

cat("\nChecking first few rows (all columns):\n")
print(head(expanded_meta))

cat("\nChecking SampleID, No_Patient, Timepoint:\n")
print(head(expanded_meta[, c("SampleID", "No_Patient", "Timepoint")]))

# Check if data is actually there
cat("\nChecking if Age column has data (first 6):\n")
print(head(expanded_meta$Age))

cat("\nChecking if Group column has data (first 6):\n")
print(head(expanded_meta$Group))
write.csv(expanded_meta, '/home/sonec_iv/RSF_res_4_ps/expanded_metadata.csv')
```
# OTU from mpa4 table
```{r}
# Load required libraries
library(phyloseq)
library(tidyverse)

# 1. Read MetaPhLAN table (preserving sample names with dashes)
# also making phyloseq w/counts
otu_table <- read.table("/home/sonec_iv/RSF_res_4_ps/mpa4/counts/merged_metaphlan_counts.txt", 
                        header = TRUE, 
                        sep = "\t", 
                        row.names = 1, 
                        check.names = FALSE, 
                        comment.char = "")
otu_table2 <- read.table("/home/sonec_iv/RSF_res_4_ps/mpa4/counts/merged_metaphlan_counts.txt", 
                        header = TRUE, 
                        sep = "\t", 
                        #row.names = 1, 
                        check.names = FALSE, 
                        comment.char = "")
#otu_table$clade_taxid  <- NULL 
# Check structure
dim(otu_table)  # Should show [taxa, samples]
head(colnames(otu_table))  # Should show sample names like "MB118-1"
otu_matrix <- as.matrix(otu_table)
otu_matrix2 <- as.matrix(otu_table2)
```

# get tax table from mpa4
```{r}
# Extract taxonomic lineages from row names
tax_strings <- otu_matrix2[,1]
#tax_strings <- rownames(otu_matrix)

# Parse the taxonomic strings
parse_metaphlan_taxonomy <- function(tax_strings) {
  # Create empty matrix
  tax_matrix <- matrix(NA, 
                       nrow = length(tax_strings), 
                       ncol = 8,
                       dimnames = list(tax_strings,
                                       c("Kingdom", "Phylum", "Class", 
                                         "Order", "Family", "Genus", "Species", "SGB")))
  
  for(i in seq_along(tax_strings)) {
    # Split by pipe
    parts <- strsplit(tax_strings[i], "\\|")[[1]]
    
    # Clean each part (remove k__, p__, etc.)
    cleaned_parts <- sub("^[a-z]__", "", parts)
    
    # Fill in available levels
    if(length(cleaned_parts) <= 8) {
      tax_matrix[i, 1:length(cleaned_parts)] <- cleaned_parts
    } else {
      # For longer lineages (some MetaPhLAN versions)
      tax_matrix[i, 1:7] <- cleaned_parts[1:7]
      tax_matrix[i, 8] <- paste(cleaned_parts[8:length(cleaned_parts)], collapse = "|")
    }
  }
  
  return(tax_matrix)
}

# Create taxonomy table
tax_matrix <- parse_metaphlan_taxonomy(tax_strings)
tax_matrix_no_sgb <- tax_matrix[, !colnames(tax_matrix) %in% "SGB"]
unique_tax <- unique(tax_matrix_no_sgb)
head(tax_matrix)
```

# WGS phyloseq

```{r}
# Create phyloseq components
expanded_meta <- read.csv('/home/sonec_iv/RSF_res_4_ps/expanded_metadata.csv',row.names = 1)
OTU <- otu_table(otu_matrix, taxa_are_rows = TRUE)
TAX <- tax_table(unique_tax)
#TAX <- tax_table(tax_matrix_no_sgb)
SAMPLES <- sample_data(expanded_meta)

# Create the phyloseq object
physeq <- phyloseq(OTU, TAX, SAMPLES)
physeq <- subset_taxa(physeq, !(Kingdom %in% c("UNCLASSIFIED","Archaea", "Eukaryota")))
saveRDS(physeq, '/home/sonec_iv/RSF_res_4_ps/RSF_all_physeq_wgs_COUNTS_v2.rds')
physeq <- readRDS('/home/sonec_iv/RSF_res_4_ps/RSF_all_physeq_wgs_COUNTS_v2.rds')
# Check the object
physeq@tax_table
cat("\nNumber of taxa:", ntaxa(physeq))
cat("\nNumber of samples:", nsamples(physeq))
cat("\nSample variables:", sample_variables(physeq))
level <- 'Species'
#2level <- 'Family'
det <- 1/100
prev <- 10/100
taxas <- core_members(physeq, detection = det, prevalence = prev, include.lowest = F)
ps_obj <- prune_taxa(taxas, physeq)
ps_obj <- speedyseq::tax_glom(ps_obj, taxrank = level, NArm = TRUE)
ps_obj@tax_table
ps_obj2 <- speedyseq::tax_glom(physeq, taxrank = level, NArm = TRUE)
ps_obj2@tax_table
```

#WGS phyloseq DEBUG
```{r}
# Load required libraries
library(phyloseq)
library(dplyr)

# Load your data (adjust file paths/loading methods as needed)
# Assuming your data is already loaded as:
# otu_table, tax_table, sample_data

# First, let's examine the structure of each input
cat("=== DATA STRUCTURE ===\n")
cat("OTU table dimensions:", dim(otu_table), "\n")
cat("TAX table dimensions:", dim(unique_tax), "\n")
cat("SAMPLES table dimensions:", dim(expanded_meta), "\n\n")

# Check row and column names
cat("=== NAMES COMPARISON ===\n")

# 1. Compare sample names between OTU table and SAMPLES table
otu_samples <- colnames(otu_table)
sample_samples <- rownames(expanded_meta)

cat("Samples in OTU table but NOT in SAMPLES table:\n")
missing_in_samples <- setdiff(otu_samples, sample_samples)
print(missing_in_samples)
cat("Count:", length(missing_in_samples), "\n\n")

cat("Samples in SAMPLES table but NOT in OTU table:\n")
missing_in_otu <- setdiff(sample_samples, otu_samples)
print(missing_in_otu)
cat("Count:", length(missing_in_otu), "\n\n")

# 2. Compare OTU/taxa names between OTU table and TAX table
otu_taxa <- rownames(otu_table)
tax_taxa <- rownames(tax_table)

cat("OTUs in OTU table but NOT in TAX table:\n")
missing_in_tax <- setdiff(otu_taxa, tax_taxa)
cat("Count:", length(missing_in_tax), "\n")
if(length(missing_in_tax) > 0) {
  cat("First few:", head(missing_in_tax), "\n")
}

cat("\nOTUs in TAX table but NOT in OTU table:\n")
missing_in_otu_rows <- setdiff(tax_taxa, otu_taxa)
cat("Count:", length(missing_in_otu_rows), "\n\n")

# 3. Check for duplicates
cat("=== DUPLICATE CHECK ===\n")
cat("Duplicate sample names in OTU table:", any(duplicated(otu_samples)), "\n")
cat("Duplicate sample names in SAMPLES table:", any(duplicated(sample_samples)), "\n")
cat("Duplicate taxa names in OTU table:", any(duplicated(otu_taxa)), "\n")
cat("Duplicate taxa names in TAX table:", any(duplicated(tax_taxa)), "\n\n")

# 4. Check for special characters or spaces in names
cat("=== NAME FORMATTING ISSUES ===\n")
cat("Sample names with spaces in OTU table:\n", otu_samples[grepl("\\s", otu_samples)], "\n")
cat("Sample names with spaces in SAMPLES table:\n", sample_samples[grepl("\\s", sample_samples)], "\n")

# 5. Create diagnostic data frames for manual inspection
comparison_df <- data.frame(
  Sample = unique(c(otu_samples, sample_samples)),
  In_OTU = unique(c(otu_samples, sample_samples)) %in% otu_samples,
  In_SAMPLES = unique(c(otu_samples, sample_samples)) %in% sample_samples
)

cat("\n=== SAMPLE OVERLAP SUMMARY ===")
cat("\nTotal unique samples across both tables:", nrow(comparison_df))
cat("\nSamples in both tables:", sum(comparison_df$In_OTU & comparison_df$In_SAMPLES))
cat("\nSamples only in OTU table:", sum(comparison_df$In_OTU & !comparison_df$In_SAMPLES))
cat("\nSamples only in SAMPLES table:", sum(!comparison_df$In_OTU & comparison_df$In_SAMPLES), "\n\n")

# 6. Try creating phyloseq object step by step
cat("=== PHYLOSEQ CREATION DEBUG ===\n")

# Create individual phyloseq components
cat("Creating OTU table for phyloseq...\n")
otu_mat <- as.matrix(otu_table)
OTU <- otu_table(otu_mat, taxa_are_rows = TRUE)

cat("Creating TAX table for phyloseq...\n")
tax_mat <- unique_tax
TAX <- tax_table(tax_mat)

cat("Creating sample data for phyloseq...\n")
sam_df <- sample_data(expanded_meta)

# Check which samples are shared between OTU and sample data
shared_samples <- intersect(otu_samples, sample_samples)
cat("\nShared samples between OTU and sample data:", length(shared_samples), "\n")

# Subset to only shared samples
if(length(shared_samples) < length(otu_samples)) {
  cat("\nSubsetting data to shared samples only...\n")
  OTU_subset <- otu_table(otu_mat[, shared_samples], taxa_are_rows = TRUE)
  SAM_subset <- sample_data(sam_df[shared_samples, ])
  
  # Try creating phyloseq with subset
  physeq_subset <- phyloseq(OTU_subset, TAX, SAM_subset)
  cat("Phyloseq object created with", nsamples(physeq_subset), "samples\n")
  
  # Compare with original attempt
  cat("\nOriginal phyloseq attempt would have", length(shared_samples), "samples\n")
  cat("Difference:", length(otu_samples) - length(shared_samples), "samples missing\n")
}

# 7. Check if the issue might be with taxa filtering
cat("\n=== TAXA FILTERING CHECK ===\n")
# Phyloseq removes taxa that don't appear in any samples after subsetting
# Let's check if any samples have zero reads after taxa filtering

if(exists("physeq_subset")) {
  sample_sums <- sample_sums(physeq_subset)
  zero_read_samples <- names(sample_sums[sample_sums == 0])
  cat("Samples with zero total reads:", length(zero_read_samples), "\n")
  if(length(zero_read_samples) > 0) {
    cat("These samples might be automatically removed by phyloseq\n")
  }
}

# 8. Print final diagnostic summary
cat("\n=== DIAGNOSTIC SUMMARY ===\n")
cat("Expected samples (from OTU table):", length(otu_samples), "\n")
cat("Expected samples (from SAMPLES table):", length(sample_samples), "\n")
cat("Shared samples:", length(intersect(otu_samples, sample_samples)), "\n")
```



# create HUMANn phyloseq object
```{r}
expanded_meta <- read.csv('/home/sonec_iv/RSF_res_4_ps/expanded_metadata.csv',row.names = 1)
#create HUMANn phyloseq
humann_data <- read.csv("/home/sonec_iv/RSF_res_4_ps/hm3/RSF_humann_res.csv",header = TRUE, 
                        sep = "\t", 
                        row.names = 1, 
                        check.names = FALSE, 
                        comment.char = "")
#humann_data <- humann_data %>%column_to_rownames(var = "Pathway")
humann_data <- humann_data[!grepl("UNINTEGRATED|UNMAPPED", rownames(humann_data)), ]
humann_data <- humann_data[!grepl("archaea|fungi|yeast|mammals|vertebrates|plants|mitochondria|invertebrates|mammalian|human", rownames(humann_data)), ]

```

#humann phyloseq DEBUG
```{r}
# Load required library
library(phyloseq)

# Read your data (adjust paths if needed)
expanded_meta <- read.csv('/home/sonec_iv/RSF_res_4_ps/expanded_metadata.csv', row.names = 1)
humann_data <- read.csv("/home/sonec_iv/RSF_res_4_ps/hm3/RSF_humann_res.csv", 
                        header = TRUE, 
                        sep = "\t", 
                        row.names = 1, 
                        check.names = FALSE, 
                        comment.char = "")

# Check initial dimensions
cat("=== INITIAL DIMENSIONS ===\n")
cat("Metadata samples:", nrow(expanded_meta), "\n")
cat("Humann data samples (columns):", ncol(humann_data), "\n\n")

# Get sample names from each source
meta_samples <- rownames(expanded_meta)
humann_samples <- colnames(humann_data)

# Find mismatches
cat("=== SAMPLE NAME MISMATCHES ===\n")

# Samples in metadata but NOT in humann data
missing_in_humann <- setdiff(meta_samples, humann_samples)
cat("Samples in metadata but NOT in humann data:", length(missing_in_humann), "\n")
if (length(missing_in_humann) > 0) {
  cat("Missing samples:\n")
  print(missing_in_humann)
}

# Samples in humann data but NOT in metadata
missing_in_meta <- setdiff(humann_samples, meta_samples)
cat("\nSamples in humann data but NOT in metadata:", length(missing_in_meta), "\n")
if (length(missing_in_meta) > 0) {
  cat("Extra samples in humann data:\n")
  print(missing_in_meta)
}

# Shared samples
shared_samples <- intersect(meta_samples, humann_samples)
cat("\nShared samples (will be in phyloseq):", length(shared_samples), "\n")

# Diagnostic summary
cat("\n=== DIAGNOSTIC SUMMARY ===\n")
cat("Total metadata samples:", length(meta_samples), "\n")
cat("Total humann data samples:", length(humann_samples), "\n")
cat("Shared samples:", length(shared_samples), "\n")
cat("Missing from phyloseq (expected):", length(meta_samples) - length(shared_samples), "\n")

# Check for common formatting issues
cat("\n=== FORMATTING CHECKS ===\n")

# Check for whitespace differences
meta_trimmed <- trimws(meta_samples)
humann_trimmed <- trimws(humann_samples)

# Find samples that match after trimming whitespace
potential_matches <- setdiff(meta_samples, meta_trimmed)
cat("Metadata samples with leading/trailing whitespace:", length(potential_matches), "\n")

potential_matches <- setdiff(humann_samples, humann_trimmed)
cat("Humann samples with leading/trailing whitespace:", length(potential_matches), "\n")

# Check for case sensitivity issues
lower_meta <- tolower(meta_samples)
lower_humann <- tolower(humann_samples)

# Find case-insensitive matches that aren't exact matches
case_insensitive_matches <- intersect(lower_meta, lower_humann)
exact_matches <- intersect(meta_samples, humann_samples)
cat("\nCase-insensitive matches:", length(case_insensitive_matches), "\n")
cat("Exact case matches:", length(exact_matches), "\n")

if (length(case_insensitive_matches) > length(exact_matches)) {
  cat("Potential case sensitivity issues detected!\n")
  
  # Find which samples differ only by case
  case_diff <- setdiff(case_insensitive_matches, tolower(exact_matches))
  cat("\nSamples that match ignoring case but not with exact case:\n")
  
  for (sample in case_diff) {
    meta_match <- meta_samples[tolower(meta_samples) == sample]
    humann_match <- humann_samples[tolower(humann_samples) == sample]
    cat("  Metadata:", meta_match, " | Humann:", humann_match, "\n")
  }
}

# Check for special characters or problematic characters
cat("\n=== SPECIAL CHARACTER CHECKS ===\n")
problem_chars <- c(" ", "-", "_", ".", ":", ";", ",", "(", ")", "[", "]", "{", "}", "/", "\\")

check_problem_chars <- function(names) {
  has_problem <- sapply(names, function(x) any(sapply(problem_chars, function(char) grepl(char, x, fixed = TRUE))))
  return(names[has_problem])
}

cat("Metadata samples with potential problem characters:\n")
print(check_problem_chars(meta_samples))

cat("\nHumann samples with potential problem characters:\n")
print(check_problem_chars(humann_samples))

# Try to create phyloseq with only shared samples to verify
cat("\n=== TESTING PHYLOSEQ CREATION ===\n")

# Create subsets with only shared samples
shared_samples <- intersect(meta_samples, humann_samples)

if (length(shared_samples) > 0) {
  # Subset both tables to shared samples
  meta_subset <- expanded_meta[shared_samples, , drop = FALSE]
  humann_subset <- humann_data[, shared_samples, drop = FALSE]
  
  # Create phyloseq object
  test_physeq <- phyloseq(
    otu_table(humann_subset, taxa_are_rows = TRUE),
    sample_data(meta_subset)
  )
  
  cat("Test phyloseq created with", nsamples(test_physeq), "samples\n")
  cat("This matches expected:", length(shared_samples) == nsamples(test_physeq), "\n")
}

# Create a comprehensive comparison table
cat("\n=== COMPREHENSIVE SAMPLE COMPARISON ===\n")
all_samples <- unique(c(meta_samples, humann_samples))
comparison_df <- data.frame(
  Sample = all_samples,
  In_Metadata = all_samples %in% meta_samples,
  In_Humann = all_samples %in% humann_samples,
  In_Both = all_samples %in% meta_samples & all_samples %in% humann_samples
)

# Sort for easier viewing
comparison_df <- comparison_df[order(comparison_df$In_Both, decreasing = TRUE), ]

# Show only mismatched samples for debugging
mismatched <- comparison_df[!comparison_df$In_Both, ]
cat("\nMISMATCHED SAMPLES (", nrow(mismatched), " total):\n", sep = "")
print(mismatched)

# Save mismatched samples to file for reference
write.csv(mismatched, "mismatched_samples_debug.csv", row.names = FALSE)
cat("\nMismatched samples saved to 'mismatched_samples_debug.csv'\n")

# Suggestions for fixing
cat("\n=== SUGGESTED FIXES ===\n")
cat("1. Check for whitespace differences:\n")
cat("   colnames(humann_data) <- trimws(colnames(humann_data))\n")
cat("   rownames(expanded_meta) <- trimws(rownames(expanded_meta))\n\n")

cat("2. Standardize naming conventions:\n")
cat("   # Remove special characters\n")
cat("   clean_names <- function(x) {\n")
cat("     x <- gsub('[^[:alnum:]]', '_', x)  # Replace non-alphanumeric with underscore\n")
cat("     x <- gsub('_+', '_', x)            # Collapse multiple underscores\n")
cat("     x <- gsub('^_|_$', '', x)          # Remove leading/trailing underscores\n")
cat("     return(x)\n")
cat("   }\n")
cat("   colnames(humann_data) <- clean_names(colnames(humann_data))\n")
cat("   rownames(expanded_meta) <- clean_names(rownames(expanded_meta))\n\n")

cat("3. Force match by subsetting to intersection:\n")
cat("   shared <- intersect(rownames(expanded_meta), colnames(humann_data))\n")
cat("   expanded_meta <- expanded_meta[shared, ]\n")
cat("   humann_data <- humann_data[, shared]\n")
cat("   # Then create phyloseq\n")
```



# making also HUMANn joint phyloseq, without splitting pathways to specific bacteria
```{r}
humann_data_joint <- humann_data[!grepl("\\|", row.names(humann_data)), ]  # Rows without '|'
hm_phyloseq <- phyloseq(otu_table(humann_data,taxa_are_rows = TRUE), sample_data(expanded_meta))
hm_phyloseq_joint <- phyloseq(otu_table(humann_data_joint,taxa_are_rows = TRUE), sample_data(expanded_meta))
#hm_phyloseq <- subset_samples(hm_phyloseq, !(ID %in% samples_2_del))
#hm_phyloseq_joint <- subset_samples(hm_phyloseq_joint, !(ID %in% samples_2_del))
saveRDS(hm_phyloseq_joint, '/home/sonec_iv/RSF_res_4_ps/hm3/RSF_ps_humann_joint.rds')
saveRDS(hm_phyloseq, '/home/sonec_iv/RSF_res_4_ps/hm3/RSF_ps_humann.rds')
```

# MAGs phyloseq and its filtering
```{r}
expanded_meta <- read.csv('/home/sonec_iv/RSF_res_4_ps/expanded_metadata.csv',row.names = 1)
mags_count_data <- read.csv("/home/sonec_iv/RSF_res_4_ps/mags/RSF2_mags_reads.csv", sep=',', check.names = FALSE)
mags_tax_data <- read.csv("/home/sonec_iv/RSF_res_4_ps/mags/RSF2_gtdbtk.tsv", sep='\t', header=FALSE)
mags_tax_data$Species <- gsub(" ", "_", mags_tax_data$Species)
mags_tax_data<- mags_tax_data %>% mutate_all(na_if,"")
tax_matrix <- as.matrix(mags_tax_data[, -1])
rownames(tax_matrix) <- mags_tax_data[, 1]    # 1st col is rownames now
colnames(tax_matrix) <- colnames(mags_tax_data)[-1]
mags_tax_table <- tax_table(tax_matrix)
mags_ps <- phyloseq(otu_table(mags_count_data,taxa_are_rows = TRUE), sample_data(meta),tax_table(mags_tax_table))
mags_ps <- subset_samples(mags_ps, !(ID %in% samples_2_del))
mags_ps@otu_table[is.na(mags_ps@otu_table)] <- 0
# filtering out of bad quality MAGs
keep_genomes <- read.csv("/home/sonec_iv/RSF_res_4_ps/mags/derep/RSF1_derep_mags.list")
keep_genomes <- as.character(keep_genomes)
keep_genomes_clean <- gsub('[\\"c()]', '', keep_genomes)  # Remove \", "c(", ")", etc.
keep_genomes_clean <- trimws(keep_genomes_clean)
keep_genomes_clean <- unlist(strsplit(keep_genomes_clean, ", "))
if (!taxa_are_rows(mags_ps)) {
  otu_table(mags_ps) <- t(otu_table(mags_ps))
}
# Subset taxa (genomes)
mags_ps_filtered <- prune_taxa(taxa_names(mags_ps) %in% keep_genomes_clean, mags_ps)
#saveRDS(mags_ps_filtered, '/home/sonec_iv/RSF_res_4_ps/RSF2_ps_mags_filt.rds')
```


# MAGs phyloseq pt.2
```{r}
mags_count_data <- read.csv("/home/sonec_iv/RSF_res_4_ps/mags/RSF1_mags_reads.csv", sep=',', check.names = FALSE)
mags_tax_data <- read.csv("/home/sonec_iv/RSF_res_4_ps/mags/RSF1_gtdbtk.tsv", sep='\t', header=FALSE, row.names=1)
#meta <- read.csv("/home/sonec_iv/RSF_res_4_ps/expanded_metadata.csv", row.names = 1)

# Create a new dataframe with split taxonomy columns
df_new <- mags_tax_data %>%
  # Separate the classification column into multiple columns
  separate(col = V2, 
           into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
           sep = ";") %>%
  # Remove the prefixes (d__, p__, etc.) from each column
  mutate(across(c(Kingdom, Phylum, Class, Order, Family, Genus, Species),
                ~ str_remove(., "^[a-z]__"))) %>%
  # Select only the first column (assuming it's an ID or key column) and the new taxonomy columns
  select(1, Kingdom, Phylum, Class, Order, Family, Genus, Species)
#rownames(df_new) <- df_new$user_genome
#df_new$user_genome <- NULL
# View the result
mags_count_data[is.na(mags_count_data)] <- 0
otu_matrix <- mags_count_data %>%
  column_to_rownames("genome") %>%  # Move taxa_id to row names
  as.matrix()
otu <- otu_table(otu_matrix, taxa_are_rows = T)
otu_taxa <- taxa_names(otu)
#tax <- tax_table(df_new)
#tax_taxa <- rownames(tax)
#otu <- otu_table(otu_matrix, taxa_are_rows = TRUE)
mags_tax_table <- tax_table(as.matrix(df_new))
mags_ps <- phyloseq(otu, sample_data(expanded_meta),mags_tax_table)
saveRDS(mags_ps,'/home/sonec_iv/RSF_res_4_ps/mags/RSF1_ps.rds')
```
# MAGs ps DEBUG
```{r}
# Display structure and parts of your dataframes
cat("=== OTU DATA (abundance) ===\n")
cat("Dimensions:", dim(otu_matrix), "\n")
cat("First few rownames of otu_data:\n", head(rownames(otu_matrix)), "\n\n")

cat("=== TAXONOMY DATA ===\n")
cat("Dimensions:", dim(df_new), "\n")
cat("First few rownames of tax_data:\n", head(rownames(df_new)), "\n\n")

# Check if rownames are identical
cat("=== COMPARING ROWNAMES ===\n")
cat("Are rownames identical?", identical(rownames(otu_matrix), rownames(df_new)), "\n")
cat("Do they have the same length?", length(rownames(otu_matrix)) == length(rownames(df_new)), "\n")
cat("First mismatched rowname (if any):\n")
mismatch <- which(rownames(otu_matrix) != rownames(df_new))
if(length(mismatch) > 0) {
  cat("At position", mismatch[1], "otu:", rownames(otu_matrix)[mismatch[1]], 
      "tax:", rownames(df_new)[mismatch[1]], "\n")
} else {
  cat("No mismatches found\n")
}
```

# DEBUG and fix
```{r}
# Step 1: Check which MAGs are in common vs unique to each dataset
common_mags <- intersect(rownames(otu_matrix), rownames(df_new))
otu_only_mags <- setdiff(rownames(otu_matrix), rownames(df_new))
tax_only_mags <- setdiff(rownames(df_new), rownames(otu_matrix))

cat("=== TAXA OVERLAP ANALYSIS ===\n")
cat("Total OTU MAGs:", length(rownames(otu_matrix)), "\n")
cat("Total TAX MAGs:", length(rownames(df_new)), "\n")
cat("Common MAGs:", length(common_mags), "\n")
cat("MAGs in OTU only:", length(otu_only_mags), "\n")
cat("MAGs in TAX only:", length(tax_only_mags), "\n\n")

# Show examples of mismatched MAGs
cat("First 5 MAGs in OTU only:\n")
print(head(otu_only_mags, 5))
cat("\nFirst 5 MAGs in TAX only:\n")
print(head(tax_only_mags, 5))

# Step 2: Subset both datasets to only include common MAGs
# This is usually the safest approach
otu_matrix_common <- otu_matrix[common_mags, ]
df_new_common <- df_new[common_mags, ]

cat("\n=== AFTER SUBSETTING TO COMMON MAGs ===\n")
cat("OTU matrix dimensions:", dim(otu_matrix_common), "\n")
cat("TAX matrix dimensions:", dim(df_new_common), "\n")

# Step 3: Check if rownames now match
cat("Rownames identical after subsetting?", identical(rownames(otu_matrix_common), rownames(df_new_common)), "\n")

# Step 4: Convert taxonomy data to matrix properly
# Make sure all columns are character type
tax_matrix <- as.matrix(df_new_common)
cat("\nTaxonomy matrix rownames (first 5):\n")
print(head(rownames(tax_matrix)))

# Step 5: Create phyloseq object
otu <- otu_table(otu_matrix_common, taxa_are_rows = TRUE)
tax <- tax_table(tax_matrix)

physeq <- phyloseq(otu, tax)

cat("\n=== PHYLOSEQ CREATION SUCCESS ===\n")
cat("Phyloseq object created!\n")
cat("Number of taxa in phyloseq:", ntaxa(physeq), "\n")
cat("Number of samples in phyloseq:", nsamples(physeq), "\n")
cat("First 5 taxa names in phyloseq:\n")
print(head(taxa_names(physeq), 5))
saveRDS(physeq,'/home/sonec_iv/RSF_res_4_ps/mags/RSF1_ps.rds')
```

# merge MAGs phyloseqs
```{r}
p1 <- readRDS('/home/sonec_iv/RSF_res_4_ps/mags/RSF1_ps.rds')
p2 <- readRDS('/home/sonec_iv/RSF_res_4_ps/mags/RSF2_ps.rds')
p3 <- readRDS('/home/sonec_iv/RSF_res_4_ps/mags/RSF3_ps.rds')
ps_mags <- merge_phyloseq(p1,p2,p3)
saveRDS(ps_mags, '/home/sonec_iv/RSF_res_4_ps/RSF_ps_mags.rds')
ps_obj2 <- ps_mags
level <- 'Species'
det <- 1/100
prev <- 10/100
taxas <- core_members(ps_obj2, detection = det, prevalence = prev, include.lowest = F)
ps_obj <- prune_taxa(taxas, ps_obj2)
ps_obj <- speedyseq::tax_glom(ps_obj, taxrank = level, NArm = TRUE)
ps_obj@tax_table
ps_mags
```

# merged MAGs phyloseq DEBUG
```{r}
# Load required library
library(phyloseq)
#ps_mags <- readRDS('/home/sonec_iv/RSF_res_4_ps/RSF_ps_mags.rds')
# Get sample names from both sources
meta_samples <- rownames(expanded_meta)
mags_samples <- sample_names(ps_mags)

# Basic dimensions
cat("=== SAMPLE COUNTS ===\n")
cat("Metadata samples:", length(meta_samples), "\n")
cat("MAGs phyloseq samples:", length(mags_samples), "\n")
cat("Missing samples from MAGs:", length(meta_samples) - length(mags_samples), "\n\n")

# Find mismatches
cat("=== SAMPLE NAME COMPARISON ===\n")

# Samples in metadata but NOT in MAGs phyloseq
missing_in_mags <- setdiff(meta_samples, mags_samples)
cat("Samples in metadata but NOT in MAGs (", length(missing_in_mags), " samples):\n", sep = "")
if (length(missing_in_mags) > 0) {
  print(missing_in_mags)
  
  # Show first 10 for brevity
  if (length(missing_in_mags) > 10) {
    cat("\nFirst 10 missing samples:\n")
    print(head(missing_in_mags, 10))
  }
}

# Samples in MAGs phyloseq but NOT in metadata
extra_in_mags <- setdiff(mags_samples, meta_samples)
cat("\nSamples in MAGs but NOT in metadata (", length(extra_in_mags), " samples):\n", sep = "")
if (length(extra_in_mags) > 0) {
  print(extra_in_mags)
} else {
  cat("None\n")
}

# Shared samples
shared_samples <- intersect(meta_samples, mags_samples)
cat("\nShared samples (present in both):", length(shared_samples), "\n")

# Diagnostic summary
cat("\n=== DIAGNOSTIC SUMMARY ===\n")
cat("Total expected (from metadata):", length(meta_samples), "\n")
cat("Present in MAGs phyloseq:", length(mags_samples), "\n")
cat("Missing from MAGs:", length(missing_in_mags), "\n")
cat("Extra in MAGs (not in metadata):", length(extra_in_mags), "\n")
cat("Overlap (shared):", length(shared_samples), "\n\n")

# Check if MAGs phyloseq might have filtered out samples with zero reads
cat("=== CHECKING FOR ZERO-READ SAMPLES ===\n")
if (any(sample_sums(ps_mags) == 0)) {
  zero_read_samples <- sample_names(ps_mags)[sample_sums(ps_mags) == 0]
  cat("Samples in MAGs with zero total reads:", length(zero_read_samples), "\n")
  cat("These would be present in MAGs but have no data:\n")
  print(zero_read_samples)
} else {
  cat("No samples in MAGs have zero reads.\n")
}

# Create a comparison table for all samples
cat("\n=== COMPREHENSIVE COMPARISON TABLE ===\n")
all_samples <- unique(c(meta_samples, mags_samples))
comparison_df <- data.frame(
  Sample = all_samples,
  In_Metadata = all_samples %in% meta_samples,
  In_MAGs = all_samples %in% mags_samples,
  Status = ifelse(all_samples %in% meta_samples & all_samples %in% mags_samples, "Shared",
                  ifelse(all_samples %in% meta_samples & !all_samples %in% mags_samples, "Only in Metadata",
                         "Only in MAGs"))
)

# Add sample sums from MAGs for context
if (exists("ps_mags") && nsamples(ps_mags) > 0) {
  mags_sample_sums <- sample_sums(ps_mags)
  comparison_df$MAGs_Read_Sum <- ifelse(comparison_df$Sample %in% names(mags_sample_sums),
                                        mags_sample_sums[comparison_df$Sample], 0)
}

# Sort for easier viewing: shared first, then only in metadata, then only in MAGs
comparison_df <- comparison_df[order(match(comparison_df$Status, c("Shared", "Only in Metadata", "Only in MAGs")),
                                     comparison_df$Sample), ]

# Show summary counts
cat("\nStatus counts:\n")
print(table(comparison_df$Status))

# Show only missing samples (Only in Metadata) for debugging
missing_df <- comparison_df[comparison_df$Status == "Only in Metadata", ]
cat("\n=== DETAILED VIEW OF MISSING SAMPLES ===")
cat("\nTotal missing from MAGs:", nrow(missing_df), "\n")
if (nrow(missing_df) > 0) {
  print(missing_df)
}

# Save detailed comparison to file
write.csv(comparison_df, "mags_vs_metadata_comparison.csv", row.names = FALSE)
cat("\nDetailed comparison saved to 'mags_vs_metadata_comparison.csv'\n")

# Check for common issues in missing sample names
cat("\n=== ANALYZING MISSING SAMPLE NAMES ===\n")
if (length(missing_in_mags) > 0) {
  # Check for pattern differences
  cat("Checking for naming pattern differences...\n")
  
  # Example: Are missing samples named differently (e.g., different prefixes)?
  # Look at first few characters to see if there's a pattern
  if (length(missing_in_mags) >= 5) {
    cat("First 5 missing samples:\n")
    print(head(missing_in_mags, 5))
    
    # Compare with first 5 MAGs samples
    cat("\nFirst 5 MAGs samples:\n")
    print(head(mags_samples, 5))
  }
  
  # Check if missing samples have special characters that might cause issues
  cat("\nChecking for special characters in missing samples:\n")
  special_chars <- c("-", "_", ".", " ", ":", ";", "(", ")", "[", "]")
  
  has_special <- sapply(missing_in_mags, function(x) {
    any(sapply(special_chars, function(char) grepl(char, x, fixed = TRUE)))
  })
  
  if (any(has_special)) {
    cat("Missing samples with special characters:\n")
    print(missing_in_mags[has_special])
  } else {
    cat("No obvious special character issues found.\n")
  }
  
  # Check if case sensitivity might be an issue
  cat("\nChecking for case sensitivity issues...\n")
  lower_meta <- tolower(meta_samples)
  lower_mags <- tolower(mags_samples)
  
  # Find samples that match when ignoring case
  case_insensitive_matches <- intersect(lower_meta, lower_mags)
  
  # Check if any missing samples match MAGs samples when ignoring case
  potential_case_issues <- missing_in_mags[tolower(missing_in_mags) %in% lower_mags]
  if (length(potential_case_issues) > 0) {
    cat("Potential case sensitivity issues (match when ignoring case):\n")
    for (sample in potential_case_issues) {
      mags_match <- mags_samples[tolower(mags_samples) == tolower(sample)]
      cat(sprintf("  Metadata: '%s' might match MAGs: '%s'\n", sample, mags_match))
    }
  } else {
    cat("No obvious case sensitivity issues.\n")
  }
}

# Suggestions for next steps
cat("\n=== SUGGESTED NEXT STEPS ===\n")
cat("1. Check if MAGs data was originally generated for only a subset of samples.\n")
cat("2. Verify if sample names in MAGs files match metadata exactly.\n")
cat("3. Consider if samples were filtered out during MAGs processing (e.g., low quality).\n")
cat("4. Check the MAGs processing pipeline logs for these missing samples.\n\n")

cat("5. To create a phyloseq with all metadata samples (including missing ones):\n")
cat("   # Add empty entries for missing samples\n")
cat("   all_samples <- rownames(expanded_meta)\n")
cat("   mags_otu <- otu_table(ps_mags)\n")
cat("   \n")
cat("   # Create empty matrix for missing samples\n")
cat("   missing_samples <- setdiff(all_samples, sample_names(ps_mags))\n")
cat("   if (length(missing_samples) > 0) {\n")
cat("     # Get taxa from MAGs\n")
cat("     taxa <- taxa_names(ps_mags)\n")
cat("     n_taxa <- length(taxa)\n")
cat("     \n")
cat("     # Create zero matrix for missing samples\n")
cat("     zero_mat <- matrix(0, nrow = n_taxa, ncol = length(missing_samples))\n")
cat("     rownames(zero_mat) <- taxa\n")
cat("     colnames(zero_mat) <- missing_samples\n")
cat("     \n")
cat("     # Merge with existing OTU table\n")
cat("     merged_otu <- cbind(as.matrix(mags_otu), zero_mat)\n")
cat("     \n")
cat("     # Create new phyloseq object\n")
cat("     new_ps <- phyloseq(\n")
cat("       otu_table(merged_otu, taxa_are_rows = TRUE),\n")
cat("       tax_table(ps_mags),\n")
cat("       sample_data(expanded_meta)\n")
cat("     )\n")
cat("   }\n")
```

#phyloseq creation for batch comparison
```{r}
meta2 <- read.csv('/home/sonec_iv/RSF_res_4_ps/meta_dups_samples_batch_check.csv',
                        header = TRUE, 
                        #sep = "\t", 
                        row.names = 1, 
                        check.names = FALSE, 
                        comment.char = "")
otu_table <- read.table("/home/sonec_iv/RSF_res_4_ps/RSF_dups_samples_batch_check/RSF_dups_samples_batch_check_mpa.out", 
                        header = TRUE, 
                        sep = "\t", 
                        row.names = 1, 
                        check.names = FALSE, 
                        comment.char = "")
otu_table2 <- read.table("/home/sonec_iv/RSF_res_4_ps/RSF_dups_samples_batch_check/RSF_dups_samples_batch_check_mpa.out", 
                        header = TRUE, 
                        sep = "\t", 
                        #row.names = 1, 
                        check.names = FALSE, 
                        comment.char = "")
#otu_table$clade_taxid  <- NULL 
# Check structure
dim(otu_table)  # Should show [taxa, samples]
head(colnames(otu_table))  # Should show sample names like "MB118-1"
otu_matrix <- as.matrix(otu_table[,-1])
otu_matrix2 <- as.matrix(otu_table2)
# Extract taxonomic lineages from row names
tax_strings <- otu_matrix2[,1]
#tax_strings <- rownames(otu_matrix)

# Parse the taxonomic strings
parse_metaphlan_taxonomy <- function(tax_strings) {
  # Create empty matrix
  tax_matrix <- matrix(NA, 
                       nrow = length(tax_strings), 
                       ncol = 8,
                       dimnames = list(tax_strings,
                                       c("Kingdom", "Phylum", "Class", 
                                         "Order", "Family", "Genus", "Species", "SGB")))
  
  for(i in seq_along(tax_strings)) {
    # Split by pipe
    parts <- strsplit(tax_strings[i], "\\|")[[1]]
    
    # Clean each part (remove k__, p__, etc.)
    cleaned_parts <- sub("^[a-z]__", "", parts)
    
    # Fill in available levels
    if(length(cleaned_parts) <= 8) {
      tax_matrix[i, 1:length(cleaned_parts)] <- cleaned_parts
    } else {
      # For longer lineages (some MetaPhLAN versions)
      tax_matrix[i, 1:7] <- cleaned_parts[1:7]
      tax_matrix[i, 8] <- paste(cleaned_parts[8:length(cleaned_parts)], collapse = "|")
    }
  }
  
  return(tax_matrix)
}

# Create taxonomy table
tax_matrix <- parse_metaphlan_taxonomy(tax_strings)
tax_matrix_no_sgb <- tax_matrix[, !colnames(tax_matrix) %in% "SGB"]
unique_tax <- unique(tax_matrix_no_sgb)
#head(tax_matrix)
OTU <- otu_table(otu_matrix, taxa_are_rows = TRUE)
TAX <- tax_table(unique_tax)
#TAX <- tax_table(tax_matrix_no_sgb)
SAMPLES <- sample_data(meta2)

# Create the phyloseq object
physeq <- phyloseq(OTU, TAX, SAMPLES)
physeq <- subset_taxa(physeq, !(Kingdom %in% c("UNCLASSIFIED","Archaea", "Eukaryota")))
saveRDS(physeq, '/home/sonec_iv/RSF_res_4_ps/RSF_dups_samples_batch_check_all_physeq_wgs.rds')
physeq <- readRDS('/home/sonec_iv/RSF_res_4_ps/RSF_dups_samples_batch_check_all_physeq_wgs.rds')
# Check the object
physeq@tax_table
cat("\nNumber of taxa:", ntaxa(physeq))
cat("\nNumber of samples:", nsamples(physeq))
cat("\nSample variables:", sample_variables(physeq))
level <- 'Species'
#2level <- 'Family'
det <- 1/100
prev <- 10/100
taxas <- core_members(physeq, detection = det, prevalence = prev, include.lowest = F)
ps_obj <- prune_taxa(taxas, physeq)
ps_obj <- speedyseq::tax_glom(ps_obj, taxrank = level, NArm = TRUE)
ps_obj@tax_table
ps_obj2 <- speedyseq::tax_glom(physeq, taxrank = level, NArm = TRUE)
ps_obj2@tax_table

```

